

## Data Science for Business

Lecture 3 notes

Gene Leynes

License: GPL v3

## Typical initialization

```{r}
##------------------------------------------------------------------------------
## INITIALIZE
##------------------------------------------------------------------------------

## Remove all objects; perform garbage collection
rm(list=ls())
gc(reset=TRUE)

## Check for dependencies
if(!"geneorama" %in% rownames(installed.packages())){
    if(!"devtools" %in% rownames(installed.packages())){
        install.packages('devtools')
        }
    devtools::install_github('geneorama/geneorama')
    }

## Load libraries
geneorama::detach_nonstandard_packages()
# geneorama::loadinstall_libraries(c("geneorama", "knitr", "caret", "gbm", 
# 								   "glmnet", "ROCR", "pROC", "plyr", "class", 
# 								   "hmeasure", "randomForest", 
# 								   "AppliedPredictiveModeling", "data.table", 
# 								   "doParallel", "e1071", "rpart"))
geneorama::loadinstall_libraries(c("geneorama", "knitr", "data.table", "ggplot2"))


opts_chunk$set(tidy=FALSE, fig.width=12)
```

Code that shows how the original data was copied from Excel...

Not the most reproducable way to do things.  Normally you should update live from a website or database, but the wikipedia tables required manual edits (and it's just an example).


```{r}
## First time... Copy the data from excel and create an RDS file
# coindat <- clipped()
# coindat
# coindat$number <- NULL
# saveRDS(coindat, "data/Presidential_coins_wikipedia.Rds")


# presdat <- clipped()
# presdat$tookofficetext <- NULL
# presdat$leftofficetext <- NULL
# presdat <- sapply(presdat, function(x)gsub("\\[.+\\]", "", x))
# presdat <- as.data.table(presdat)
# str(presdat)
# presdat[,no:=as.integer(no)]
# presdat[,tookoffice:=as.IDate(tookoffice, "%B %d, %Y")]
# presdat[,leftoffice:=as.IDate(leftoffice, "%B %d, %Y")]
# setnames(presdat, "no", "number")
# saveRDS(presdat, "data/Presidential_office_data.Rds")

```


## Read in the saved data and fix it up

```{r}
rm(list=ls())
ls()

coindat <- readRDS("data/Presidential_coins_wikipedia.Rds")
presdat <- readRDS("data/Presidential_office_data.Rds")
ls()

str(coindat)
str(presdat)

coindat <- as.data.table(coindat)
coindat[,total := denver + philadelphia]

## Convert release date to iDate
coindat[,releasedate:=as.IDate(releasedate,  "%B %d, %Y")]
```

## Take a look at the coin data
```{r}
str(dat)

plot(total~releasedate, coindat[,list(total=total/1000, releasedate)], type="o", 
     main = paste("US President Mintage Output", "in 000's", sep="\n"))
addbg()

text(total/1000 ~ releasedate, labels = president, data = coindat)
# text(x = coindat$releasedate, y = coindat$total/1000, labels = coindat$president)
```


```{r}
## Can we merge on president?
inin(coindat$president, presdat$president)
## Looks like it... but there's a problem: Grover Cleveland is a duplicate
# merge(coindat, presdat, by = "president")

## We can merge on "Number", if we change "release" to "number"
setnames(coindat, "release", "number")

## Merge
## This isn't a perfect merge... (makes two copies of "president")
# merge(coindat, presdat, by = "number")

## This merge works:
dat <- merge(x = coindat, 
             y = presdat[i = TRUE,
                         j = list(number,
                                  tookoffice, 
                                  leftoffice, 
                                  party = as.factor(party))], 
             by = "number")
str(dat)
```

## Motivating example for data table

Let's say we want to calculate how long each president was in office

```{r}

## How should we calculate the time difference?
dat[, leftoffice - tookoffice]
dat[, (leftoffice - tookoffice) / 365]
dat[, round((leftoffice - tookoffice) / 365, 2)]
dat[, as.numeric(round((leftoffice - tookoffice) / 365, 2))]

## Better yet... we could do it right in the merge and eliminate the need
## for all these steps (now that we have it figured out)
dat <- merge(
    x = coindat, 
    y = presdat[
        i = number %in% 1:32, ## Maybe we're unsure of the join mechanics
        j = list(number,
                 tookoffice, 
                 leftoffice, 
                 party = as.factor(party),
                 yrs_in_office = as.numeric(round((leftoffice - tookoffice) / 
                                                      365.25, 2)))], 
    by = "number")
dat

```

# GGPLOT2

## Simple example

```{r}
ggplot(dat) + aes(x = releasedate, y = total) + geom_point()
```

## adding a line

```{r}
p <- ggplot(dat) + aes(x = releasedate, y = total) + geom_point()
p
p + geom_line()
```

### Generally speaking, don't make ggplot2 objects
It's good to know about, but usually causes more trouble than it's worth.
```{r}
rm(p)
```


## Closer to Deluxe

```{r}
ggplot(dat) + aes(x = releasedate, y = total, color = party) + 
    geom_line(size = 2) +
    geom_point(aes(size = yrs_in_office), color = "black")  
```

```{r}
ggplot(dat) + aes(x = releasedate, y = total) + 
    geom_line(size = 1) +
    geom_point(aes(size = yrs_in_office, color = party))  
```

```{r}
mycolors <- c('blue','purple','darkgreen','black','red','white','pink')

ggplot(dat) + aes(x = releasedate, y = total/1000) + 
    geom_line(size = 1) +
    geom_point(aes(size = yrs_in_office, color = party)) +
    xlab("Release Date") +
    ylab("Mintage (in thousands)") +
    theme(plot.title = element_text(size = 20)) +
    labs(title='President Coin Volume by date \n (note the newline --->)\n') +
    scale_colour_manual(values = mycolors) +
    expand_limits(y=-5e4) +
    theme(panel.background = element_rect(fill = "gray60", colour = "black")) +
    theme(panel.grid.major = element_line(colour = "gray40")) +
    theme(panel.grid.minor = element_line(colour = "gray70", linetype = "dotted")) + 
    geom_hline(h=0, size=2)
```

```{r}
mycolors <- c('blue','purple','darkgreen','black','red','orange','maroon')

ggplot(dat) + aes(x = tookoffice, y = total/1000) + 
    geom_line(size = 1) +
    geom_point(aes(size = yrs_in_office, color = party)) +
    xlab("Date in office") +
    ylab("Mintage (in thousands)") +
    theme(plot.title = element_text(size = 20)) +
    labs(title='President Coin Volume by \n date they were in office\n') +
    scale_colour_manual(values = mycolors) 
```


## Quick intro into reshaping

```{r}
library(reshape2)

dat_by_mint <- melt(dat,
                    id.vars=c("president", "releasedate"),
                    measure.vars = c("denver", "philadelphia"))
dat_by_mint <- as.data.table(dat_by_mint)
setnames(dat_by_mint, "variable", "mint")
setnames(dat_by_mint, "value", "volume")
dat_by_mint


ggplot(dat_by_mint) + aes(x = releasedate, y = volume/1000, color=mint) + 
    geom_line(size = 2) +
    geom_point(aes(size = 2, color = mint)) +
    ylab("Mintage (in thousands)") +
    theme(plot.title = element_text(size = 20)) +
    labs(title='President Coin Volume by mint\n')
```




